---
title: Advent of Code 2021
subtitle: Joint discussion
author: "David Selby & Heather Turner"
institute: "R-thritis & Warwick RUG"
date: "3 December 2021"
output:
  xaringan::moon_reader:
    df_print: tibble
    css: rth.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: center, middle, inverse

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE )
knitr::opts_chunk$set(comment = '#', message = FALSE)
```

```{r xaringan-logo, echo=FALSE}
if (!require(xaringanExtra)) remotes::install_github('gadenbuie/xaringanExtra')
xaringanExtra::use_logo(
  image_url = 'https://www.cfe.manchester.ac.uk/media/mhs/bmh-faculty/images/CfE-logo-150px.jpg',
  width = '67px', height = '50px',
  position = xaringanExtra::css_position(bottom = '.3em', left = '0.3em')
)
```

# Day 1: Sonar Sweep

<https://adventofcode.com/2021/day/1>

---

## Day 1 - Part 1

How many measurements are larger than the previous measurement?

```
199  (N/A - no previous measurement)
200  (increased)
208  (increased)
210  (increased)
200  (decreased)
207  (increased)
240  (increased)
269  (increased)
260  (decreased)
263  (increased)
```

---

## Day 1 - Part 2

How many **sliding three-measurement sums** are larger than the previous sum?

```
199  A           607  (N/A - no previous sum)
200  A B         618  (increased)
208  A B C       618  (no change)
210    B C D     617  (decreased)
200  E   C D     647  (increased)
207  E F   D     716  (increased)
240  E F G       769  (increased)
269    F G H     792  (increased)
260      G H
263        H
```

---

class: center, middle

## Day 1 - Solutions

---


### Day 1: How to make life harder for yourself

By Claire Little
.pull-right[.red[10,000 μs]]

```r
library(tidyverse)
# Part 1:
input %>% mutate(change = V1 - lag(V1),
                 sign = change / abs(change)) %>%
  group_by(sign) %>%
  summarise(n = n())

# Part 2:
input %>% mutate(window = V1 + lag(V1) + lag(V1, 2),
                 change = window - lag(window),
                 sign = change / abs(change)) %>%
  group_by(sign) %>%
  summarise(n = n())
```

???

Day 1 was well easy but I clearly need to get better at "I wonder if there's a function for that" rather than trying to bosh something together i.e. `change/abs(change)`

Would have tried to do it without tidyverse if I had more time but it was just calling out for some tidyverseness

---

### Day 1: base R using `diff()`

By David Selby
.pull-right[.red[~100 μs]]

```r
increases <- function(depths) {
  sum(diff(depths) > 0)
}
  
rolling_sum <- function(depths) {
  head(depths, -2) + tail(depths, -2) +
    head(tail(depths, -1), -1)
}
  
input1 <- scan('input01.txt')
increases(input1)
increases(rolling_sum(input1))
```

.footnote[Or you could use `zoo::rollsum(depths, 3)`]

???

100 μs is around 100 times faster than 10 ms (= 10,000 μs)

---

### Day 1: one-liner using `diff(lag = n)`

By David Schoch
.pull-right[.red[~100 μs]]

```r
sapply(c(1, 3), function(n) sum(diff(input1, lag = n) > 0))
```

--

Or with R's new anonymous function `\(x)` syntax:

```r
sapply(c(1, 3), \(n) sum(diff(input1, n) > 0))
```

???

100 μs is around 100 times faster than 10 ms (= 10,000 μs)

Worth pointing out that you can't attempt part 2 until you've completed and submitted part 1, so nobody is going to have a one-liner as their first solution!

---

class: center, middle, inverse

# Day 2: Dive!

<https://adventofcode.com/2021/day/2>

---

## Day 2 - Part 1

- `forward X` increases the horizontal position by `X` units.
- `down X` increases the depth by `X` units.
- `up X` decreases the depth by `X` units.

```
                  x  depth
forward 5   -->   5      -
down 5      -->          5
forward 8   -->  13
up 3        -->          2
down 8      -->         10
forward 2   -->  15

==> x = 15, depth = 10
```

---

## Day 2 - Part 2

- `down X` increases your aim by `X` units.
- `up X` decreases your aim by `X` units.
- `forward X` does two things:
    - It increases your horizontal position by `X` units.
    - It increases your depth by your aim **multiplied by** `X`.
    
```
                  x  aim  depth
forward 5   -->   5    -      - 
down 5      -->        5      
forward 8   -->  13          40
up 3        -->        2
down 8      -->       10
forward 2   -->  15          60

==> x = 15, depth = 60
```

---

class: center, middle

## Day 2 - Solutions

---

### Day 2: I don't care if loops are inefficient

By Claire Little
.pull-right[.red[60,000 μs]]

```r
inst = read.table("path/to/Day2.csv")
ninst = dim(inst)[1]

# Day 2 - Part 1
position = c(0, 0)

for (i in 1:ninst) {
  if (inst[i, 1] == 'forward') {
    position = position + c(inst[i, 2], 0)
  } else if (inst[i, 1] == 'down') {
    position = position + c(0, inst[i, 2])
  } else {
    position = position - c(0, inst[i, 2])
  }
}
  
answer1 = position[1] * position[2]
```

???

Day 2 - I forgot how many loops & if statements I used last year and I liked using them but also appreciate it wouldn't have worked very well with bigger datasets. But it was efficient enough for what we needed

---

### Day 2: I don't care if loops are inefficient

By Claire Little
.pull-right[.red[60,000 μs]]

```r
# Day 2 - Part 2
position = c(0, 0, 0)

for (i in 1:ninst) {
  if (inst[i, 1] == 'forward') {
    position = position + c(inst[i, 2], 0, 0)
    position = position + c(0, (inst[i, 2] * position[3]), 0)
  } else if (inst[i, 1] == 'down') {
    position = position + c(0, 0, inst[i, 2])
  } else {
    position = position - c(0, 0, inst[i, 2])
  }
}

answer2 = position[1] * position[2]
```

???

Day 2 - I forgot how many loops & if statements I used last year and I liked using them but also appreciate it wouldn't have worked very well with bigger datasets. But it was efficient enough for what we needed

---

### Day 2: no ifs, no loops

By David Selby
.pull-right[.red[100 μs]]

```r
course <- read.table('input.txt', col.names = c('dir', 'value'))

with(course, {
  x <- (dir == 'forward') * value
  y <- ((dir == 'down') - (dir == 'up')) * value
  sum(x) * sum(y)
})
```

```r
with(course, {
  x <- (dir == 'forward') * value
  y <- ((dir == 'down') - (dir == 'up')) * value
  depth <- cumsum(y) * x
  sum(x) * sum(depth)
})
```

???

100 μs is around **600 times faster** than 60 ms (= 60,000 μs)

---

### Day 2: tidyverse-style

By David Selby
.pull-right[.red[3000 μs]]

```r
course %>%
  mutate(x = if_else(dir == 'forward', value, 0L),
         y = if_else(dir == 'down', value, -value) * !x,
       depth = cumsum(y) * x) %>%
  summarise(part1 = sum(x) * sum(y),
            part2 = sum(x) * sum(depth))
```

???

At 3 ms (= 3000 μs), this tidyverse solution is around 30 times slower than the base R (100 μs), but 20 times faster than loops (60,000 μs / 60 ms)

---

layout: false
class: center, middle

# Thanks!

`david.selby@manchester.ac.uk`

.small[For these slides: search `"rthritis" manchester`]

---

class: center, middle

## Next meeting

Friday 14 January 2022

.big[Writing your own R package]

_To be confirmed_
