---
title: "Lecture 10: Survival Analysis"
description: |
  Censoring. Survival curves and life tables. Comparing survival curves. Parametric regression. Cox regression.
author:
  - name: David Selby
    url: https://www.research.manchester.ac.uk/portal/david.selby.html
    affiliation: Centre for Epidemiology Versus Arthritis
    affiliation_url: https://www.cfe.manchester.ac.uk
    orcid_id: 0000-0001-8026-5663
date: 2021-02-01
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 2
preview: survival_files/figure-html5/ggsurv-confint-1.png
---

This worksheet is based on the tenth lecture of the [**Statistical Modelling in Stata**](https://personalpages.manchester.ac.uk/staff/mark.lunt/stats_course.html) course, created by Dr Mark Lunt and offered by the [Centre for Epidemiology Versus Arthritis](https://www.cfe.manchester.ac.uk/) at the University of Manchester.
The original Stata exercises and solutions are here translated into their R equivalents.

Refer to the original slides and Stata worksheets [here](https://personalpages.manchester.ac.uk/staff/mark.lunt/stats_course.html).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 6, fig.height = 5)
basedir <- 'http://personalpages.manchester.ac.uk/staff/mark.lunt'
datadir <- file.path(basedir, 'stats/10_survival/data')
```

## Life tables and survival curves

This section uses the dataset [`leukaemia.csv`](leukaemia.csv).

```{r load-leukaemia, echo = -1}
if (!file.exists('leukaemia.csv')) {
  leukaemia <- foreign::read.dta(file.path(datadir, 'leukaemia.dta'))
  leukaemia$relapse <- as.integer(leukaemia$relapse == '1:yes')
  write.csv(leukaemia, 'leukaemia.csv', row.names = FALSE)
}
leukaemia <- read.csv('leukaemia.csv')
```

First, load the **survival** package.

```{r}
library(survival)
```

The time variable is `weeks`, the number of weeks to relapse.
The outcome variable is `relapse`, which is 1 if the subject had a relapse at that time and 0 if they did not.
Create a corresponding survival object using the syntax `Surv(time, status)`.
This will be on the left hand side of your model formula.

First reorder the levels of `wbc3cat` so that 'Normal' is the reference group (for ease of comparison with the Stata solutions sheet).
Similarly we can relevel `treatment1` so that 'Standard' is the first level (as otherwise 'Drug A' comes first, alphabetically).

```{r}
leukaemia <- transform(
  leukaemia,
  treatment1 = relevel(as.factor(treatment1), 'Standard'),
  treatment2 = relevel(as.factor(treatment2), 'Standard'),
  wbc3cat = factor(wbc3cat, levels = c('Normal', 'Moderate', 'High'))
)
```

### Obtain a life table for the subjects on Drug A. What is the median survival in this group (at what time does the survival function reach 0.5)?

```{r}
survA <- survfit(Surv(weeks, relapse) ~ 1,
                 data = leukaemia,
                 subset = treatment1 == 'Drug A')
summary(survA, times = unique(leukaemia$weeks)) # `times` is optional
```
At 23 weeks, the survival function drops below 0.5.

### How many subjects were lost to followup in this treatment arm?

```{r}
table(leukaemia$relapse, leukaemia$treatment1)
```

12 subjects were lost to followup on Drug A.

### Obtain a life table for the subjects on standard treatment. What is the median survival in this group?

```{r}
survStd <- update(survA, subset = treatment1 == 'Standard')
summary(survStd)
```
The survival function drops below 0.5 after 8 weeks.

### How many subjects were lost to followup in this treatment arm?

Zero (see `table` output above).

### Do the answers to your previous questions suggest that Drug A is better, worse, or the same as standard treatment?

Median survival before relapse is better on Drug A (23 weeks) than standard treatment (8 weeks).

### Produce a Kaplan-Meier curve for each of the treatments. Does this confirm your answer to the previous question?

There are some basic plotting functions in the **survival** package.

```{r survplot}
surv_t1 <- survfit(Surv(weeks, relapse) ~ treatment1, data = leukaemia)

plot(surv_t1, xlab = 'weeks', ylab = 'survival',
     col = c(2, 4), main = 'Kaplan-Meier survival estimates')

legend('topright',
       names(surv_t1$strata),
       col = c(2, 4), lty = 1,
       bty = 'n')

abline(h = 0.5, lty = 2) # median line
```

More tools for visualisation via **ggplot2** are provided in the [**survminer** package](https://rpkgs.datanovia.com/survminer/index.html), for which a helpful [cheat sheet](https://rpkgs.datanovia.com/survminer/survminer_cheatsheet.pdf) is available.
There's less chance of mixing up the strata, as a legend is generated automatically by default:

```{r ggsurvplot}
library(survminer)
ggsurvplot(surv_t1, surv.median.line = 'hv')
```

### Add a horizontal line to the graph at y = 0.5. This line represents half of the group surviving and half having a relapse: the point where it crosses the two survival curves should give you the median survival times you calculated in earlier questions.

In base R graphics you can use `abline` and in **ggplot2** there is `geom_abline` or `geom_hline`.
The `ggsurvplot` function has a dedicated argument called [`surv.median.line`](https://github.com/kassambara/survminer/issues/61) for this purpose.
See above.

### Add the number of subjects lost to followup in the two treatment arms to the corresponding time points on the graph.

This is slightly contrived because R doesn't have a direct equivalent to the `lost` command.
So here we'll show what's offered in **survival** and **survminer**.
See the [survminer vignette](https://rpkgs.datanovia.com/survminer/) for more tips.

```{r ggsurv-censor}
ggsurvplot(surv_t1,
           surv.median.line = 'hv',
           censor = TRUE,
           cumcensor = TRUE,
           ncensor.plot = TRUE)
```

In base R, use the generic plotting functions `text()` or `points()` and extract the coordinates and values from the `survfit` object.

```{r survplot-censor}
# Original plot:
plot(surv_t1, xlab = 'weeks', ylab = 'survival',
     col = c(2, 4), main = 'Kaplan-Meier survival estimates')
legend('topright', names(surv_t1$strata),
       col = c(2, 4), lty = 1, bty = 'n')
abline(h = 0.5, lty = 2)

# Add censor labels:
text(surv_t1$time, surv_t1$surv,
     labels = surv_t1$n.censor,
     # make label transparent if n.censor == 0
     col = rgb(0, 0, 0, 0:1)[1 + (surv_t1$n.censor > 0)])
```

### Add confidence bands to your plot(s).

Look in the documentation, depending on the plotting method you are using.
Both base `plot.survfit` and `survminer::ggsurv` offer arguments `conf.int = TRUE`:

```{r survfit-confint, echo = FALSE}
plot(surv_t1, xlab = 'weeks', ylab = 'survival',
     col = c(2, 4), main = 'Kaplan-Meier survival estimates',
     conf.int = TRUE)
legend('topright', names(surv_t1$strata),
       col = c(2, 4), lty = 1, bty = 'n')
```
```{r ggsurv-confint, echo = FALSE}
ggsurvplot(surv_t1,
           surv.median.line = 'hv',
           conf.int = TRUE,
           censor = TRUE)
```

Why do the confidence bands get wider over time?
Because they are based on smaller numbers.

### Use `survdiff` to perform a logrank test to compare the survival on Drug A to that on standard treatment. Is the difference between Drug A and standard treatment statistically significant?

You can actually just do this test and add it to your **survminer** plot with the arguments `pval` and `pval.method` both set to `TRUE`:

```{r}
ggsurvplot(surv_t1,
           conf.int = TRUE,
           censor = TRUE,
           pval = TRUE, pval.method = TRUE)
```
Or do it the old-fashioned way using `survdiff()`:

```{r}
survdiff(Surv(weeks, relapse) ~ treatment1, data = leukaemia)
```
The difference is statistically significant: there are fewer relapses on Drug A than expected.

### Would you have had the same answer to the previous question if you had used a Wilcoxon test in place of a logrank test?

```{r}
survdiff(Surv(weeks, relapse) ~ treatment1, data = leukaemia,
         rho = 1)
```
## Cox regression

### Have a look at the survival curves by white blood cell count. Does the white blood cell count affect survival?

```{r}
surv_wbc <- survfit(Surv(weeks, relapse) ~ wbc3cat, data = leukaemia)
ggsurvplot(surv_wbc, conf.int = TRUE)
```

### Do a cross-tabulation of `treatment1` against `wbc3cat`. Are the proportions of subjects in each of the white blood cell counts categories the same in the two treatment arms?

```{r}
library(magrittr)
xtabs(~ treatment1 + wbc3cat, data = leukaemia) %T>%
  print %>%
  proportions(margin = 1)
```
The standard treatment appears to have a greater proportion of people with high white blood cell counts, compared to those on Drug A.

*Tip.* The `%T>%` pipe from **magrittr** lets us print _and_ pass the result of `xtabs` to the next function, without having to save the table to a temporary variable.

### Given that proportion of subjects in the "High" cell count group is greater in the standard treatment arm than in the Drug A arm, would you expect this to have increased or decreased survival in this arm of the trial?

Decreased, because leukaemia is a disease that causes high numbers of abnormal blood cells.

### White blood cell count is a potential confounder, so we need to adjust for it. First, we will perform an unadjusted Cox regression to obtain the hazard ratio before adjusting. (Use function `coxph()`.) What is the hazard ratio for Drug A, and its 95% confidence interval?

```{r}
cox_unadj <- coxph(Surv(weeks, relapse) ~ treatment1, data = leukaemia)
summary(cox_unadj)
```
The hazard ratio and confidence interval are given by

```{r}
exp(cbind(coef(cox_unadj), confint(cox_unadj)))
```

### Now obtain the adjusted hazard ratio. What is the adjusted hazard ratio and its 95% confidence interval?

Fit the adjusted model and extract the hazard ratio and its 95% confidence interval:

```{r}
cox_adj <- update(cox_unadj, ~ . + wbc3cat)
exp(cbind(HR = coef(cox_adj), confint(cox_adj)))[1, ]
```

### How did the confounding by white blood cell count affect the apparent effect of Drug A? Is this what you expected from the earlier questions?

The beneficial effect of Drug A was exaggerated by the difference in white blood cell counts between the groups.
The hazard ratio is larger (i.e. chance of relapse increased) after adjusting for it.

### Now we need to test the proportional hazards assumption. First for treatment: produce a plot of the observed and predicted Kaplan Meier plots. Are the observed and predicted curves close to each other?

To check the proportional hazards assumption, we can use `cox.zph()` from the **survival** package or `ggcoxzph()` from the **survminer** package.
Check the documentation also for information on `ggcoxdiagnostics()` and `ggadjustedcurves()`.

For example:

```{r coxph}
plot(cox.zph(cox_unadj))
ggcoxdiagnostics(cox_unadj, 'schoenfeld', ox.scale = 'time')
```

We can also plot the observed versus expected survival curves, using `survfit`.

```{r}
dummy <- list(treatment1 = levels(leukaemia$treatment1))
cox_surv <- survfit(cox_unadj, newdata = dummy)

plot(surv_t1, col = c(2, 4), lty = 1,
     xlab = 'weeks', ylab = 'survival')
lines(cox_surv, col = c(2, 4), lty = 2)
legend('topright',
       c('observed', 'expected', levels(leukaemia$treatment1)),
       lty = c(1, 2, 4, 4), col = c(1, 1, 2, 4))
```

(If there is a **ggplot2** or **survminer** approach, then it is left as an exercise.)

### Now we can test the same assumption for the effect of white blood cell count. Are the observed and predicted curves close to each other?

```{r}
cox_wbc <- update(cox_unadj, ~ wbc3cat)
plot(cox.zph(cox_wbc))

dummy <- list(wbc3cat = levels(leukaemia$wbc3cat))
cox_surv <- survfit(cox_wbc, newdata = dummy)

plot(surv_wbc, col = c(2:4), lty = 1,
     xlab = 'weeks', ylab = 'survival')
lines(cox_surv, col = c(2:4), lty = 2)
legend('topright',
       c('observed', 'expected', levels(leukaemia$wbc3cat)),
       lty = c(1, 2, 4, 4, 4), col = c(1, 1, 2:4))
```

### Perform a test of proportionality using Schoenfeld residuals. Is the regression model valid?

*Note.* You can regenerate both Schoenfeld residual plots at once with

```{r, eval = F}
plot(cox.zph(cox_adj)) # not shown
```
To perform the test, simply print

```{r}
cox.zph(cox_adj)
```

None of the test statistics is significant, thereby implying that the proportional hazards assumption holds for each of the variables.

### Obtain tests of proportionality for each individual variable. Is there any evidence of non-proportional hazards?

The `cox.zph()` function did both sets of tests at once, so the answer is already given above.

## Non-proportional hazards

### There is a second drug used in this trial, stored in `treatment2`. Compare the survival curves for Drug B and standard treatment. How does the survival on Drug B compare to that on standard treatment during the first 10 weeks?

```{r surv-t2}
surv_t2 <- survfit(Surv(weeks, relapse) ~ treatment2, data = leukaemia)
plot(surv_t2, col = c(2, 4), mark.time = TRUE)
legend('topright', lty = 1, col = c(2, 4), names(surv_t2$strata))
```

```{r gg-surv-t2}
ggsurvplot(surv_t2)
```

In the first 10 weeks, Drug B has worse survival than the standard treatment.

### How does the survival on Drug B compare to that on standard treatment after the first 10 weeks?

After 10 weeks, survival on Drug B is better than the standard treatment.

### Superimpose the predicted survival curves from a Cox regression model. How do the predicted and observed curves differ?

```{r obs-vs-exp-t2}
cox_t2 <- coxph(Surv(weeks, relapse) ~ treatment2, data = leukaemia)
dummy <- list(treatment2 = levels(leukaemia$treatment2))
cox_surv2 <- survfit(cox_t2, newdata = dummy)
plot(surv_t2, col = c(2, 4))
lines(cox_surv2, col = c(2, 4), lty = 4, lwd = 2)
legend('topright', lty = 1, col = c(2, 4), names(surv_t2$strata))
legend('bottomleft', lty = c(1, 4), lwd = 1:2, c('observed', 'predicted'))
```
The observed survival curves intersect each other but the predicted curves do not (obviously, because they are based on a proportional hazards assumption).

### Perform a Cox regression of `treatment2` and `wbc3cat`. Does Drug B have a significant effect on survival?

```{r}
cox_adj2 <- coxph(Surv(weeks, relapse) ~ treatment2 + wbc3cat,
                  data = leukaemia)
exp(confint(cox_adj2))
```
The effect of Drug B on survival is not statistically significant (the hazard ratio's 95% confidence interval contains 1).

### Test the proportional hazards assumption using the Schoenfeld residuals.

```{r}
cox.zph(cox_adj2)
```

The proportional hazards assumption does not hold overall; specifically not for treatment 2.

If you prefer visual tests:

```{r}
plot(cox.zph(cox_adj2)[1])
```

###  The Kaplan-Meier curves suggest that Drug B has a negative effect on survival initially, then becomes positive. So we will test for different effects before and after 10 weeks. Generate a life table.

```{r}
summary(surv_t2)
```

### Now, for each subject followed for more than 10 weeks, we will split the data into 2 observations, one for the time up to 10 weeks and one for the time after.

There is a good tutorial [here](http://www.drizopoulos.com/courses/emc/ep03_%20survival%20analysis%20in%20r%20companion#step-functions-for-time-varying-coefficients) explaining the approach of using step functions for time-varying coefficients.
We can use the function `survSplit()` for this.

```{r}
leuk2 <- survSplit(Surv(weeks, relapse) ~ treatment2 + wbc3cat,
                   data = leukaemia, cut = 10, episode = 'time_group')
```

### Has the life table changed?

```{r}
surv_t2s <- survfit(Surv(tstart, weeks, relapse) ~ treatment2, data = leuk2)
summary(surv_t2s)
```

No.

### Examine the data. You should see that for subjects who were followed up for less than ten weeks, there is a single record. However, for those followed up for more than 10 weeks, there are two records: one with `tstart == 0` and one with `tstart == 10`.

Look at `leuk2` (or whatever you decided to call the output of `survSplit`).

### Now we can generate separate treatment variables for the treatment effect before and after 10 weeks.

This has already been done for us with the `episode` argument of `survSplit`.
We have a variable called `time_group`.

### Now fit the Cox regression model with these predictors. What is the hazard ratio for [drug B before 10 weeks] with its 95% confidence interval?

This involves an interaction between the `time_group` strata and `treatment2`.
(Explained [here](http://www.drizopoulos.com/courses/emc/ep03_%20survival%20analysis%20in%20r%20companion#step-functions-for-time-varying-coefficients).)

```{r, tidy = TRUE}
cox_split <- coxph(Surv(tstart, weeks, relapse) ~ wbc3cat + treatment2 * strata(time_group), data = leuk2)

exp(cbind(HR = coef(cox_split), confint(cox_split)))
```

### What is the hazard ratio for `t2`?

See above.

### Do these hazard ratios confirm what you were expecting?

Yes, that Drug B increased risk initially, then reduced it.

### Now test the proportional hazards assumption, using Schoenfeld residuals. Is the model now appropriate?

```{r out.width = '33%'}
cox.zph(cox_split) %T>% plot
```
The model is now appropriate and none of the predictors appear to show non-proportionality.
