---
title: "Bayesian modelling with Stan"
description: |
  Introduction to Stan, rstan and brms for multi-level Bayesian regession modelling.
author:
  - name: Belay Birlie Yimer
    url: https://www.research.manchester.ac.uk/portal/david.selby.html
    affiliation: Centre for Epidemiology Versus Arthritis
    affiliation_url: https://www.cfe.manchester.ac.uk
    orcid_id: 0000-0001-8621-6539
date: 2021-03-26
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
rinline <- function(code) {
  # Print inline R code with backticks without evaluating it
  sprintf('``` `r %s` ```', code)
}
```

## Resources

1. [**View the talk slides**](Bayesian-modelling-with-Stan.html) ([source code](Bayesian modelling with Stan.Rmd))
2. [Stan web site](https://mc-stan.org/)
3. [JAGS](https://mcmc-jags.sourceforge.io/) (Just Another Gibbs Sampler)
4. [_Statistical Rethinking_](https://xcelab.net/rm/statistical-rethinking/) book by Richard McElreath
5. Sleep deprivation study
    - [original paper](https://onlinelibrary.wiley.com/doi/full/10.1046/j.1365-2869.2003.00337.x)
    - access the dataset in R with `data(sleepstudy, package = 'lme4')`
6. [**brms**](https://paul-buerkner.github.io/brms/): Bayesian regression modelling with Stan

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/qV6Wc_f1Cgo" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### Final linear mixed model

```{r brms-fit, echo = FALSE, results = 'hide', message = FALSE}
library(brms)
fit <- brm(Reaction ~ Days + (Days | Subject),
           data = lme4::sleepstudy, verbose = FALSE,
           file = 'sleepbrm')
```

```{r, brms, eval = FALSE, echo = TRUE}
fit <- brm(Reaction ~ Days + (Days | Subject),
           data = lme4::sleepstudy)
```

```{r summary, echo=FALSE}
fit
```

```{r traceplots, echo = TRUE}
plot(fit, N = 3)
```

```{r pp, echo = TRUE}
pp_check(fit)
```


